---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `brfss2013`. Delete this note when before you submit 
your work. 

```{r load-data}
load("brfss2013.RData")
```



* * *

## Part 1: Data

The BRFSS data contains preventive health practices and risk behaviors collected from non-institutionalized adult population, aged 18 years or older via telephonics interviews.
It is a collabortaive effort between all of the states in the United States (US) and the Centers for Disease Control and Prevention (CDC).

## Sample

1. The states use a standardized core questionnaire, optional modules, and state-added questions.
2. The survey is conducted using Random Digit Dialing (RDD) techniques on both landlines and cell phones.

## Inference

1. It is an observational study, not experimental, since we are monitoring what occurs, while experiments require the primary explanatory variable in a study be assigned for each subject by the researchers.
2. It is a randomized study using Stratified Sampling. (Point #2 above)

## Generalizability

1. The study is generalizable to all non-institutionalized adults equal to or above 18 years in the US.

## Causality

1. Observational nature of the study makes the results non-causal, i.e., we can show correlation or association only.
2. Labeling variables as explanatory and response does not guarantee the relationship between the two is actually causal, even if there is an association identified between the two variables.


* * *

## Part 2: Research questions

**Research quesion 1:**
How does sleep time affect Physical and Mental Health

physhlth: Number Of Days Physical Health Not Good
menthlth: Number Of Days Mental Health Not Good
sleptim1: How Much Time Do You Sleep

Does lower sleep time indicate poor health?
We will check the correlation between Sleep Time and Health of an individual.

**Research quesion 2:**
How does Education level associate to Income and general health

educa: Education Level
income2: Income Level
genhlth: General Health

We would expect a higher education to yield a higher income, i.e. a positive correlation.
Are richer people more healthy?

**Research quesion 3:**
Does better general health correlate to longer Length Of Time Since Last Routine Checkup or 

genhlth: General Health
checkup1: Length Of Time Since Last Routine Checkup



* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research quesion 1:**

```{r}

```



**Research quesion 2:**

```{r}

```



**Research quesion 3:**

```{r}

```

